---
title: "Socio-Economic Country Clustering for Aid Prioritization"
subtitle: "A Professional Unsupervised Machine Learning Analysis using PCA, K-means, Hierarchical Clustering, and Need-Based Ranking"
author: "Eliya Christopher Nandi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
params:
  data_path: "Country-data.csv"
  k_clusters: 4
  run_autoencoder: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10.5,
  fig.height = 6.4,
  out.width = "96%",
  comment = "#>"
)

set.seed(123)
```

<style>
/* ===============================
   Professional Report Styling
   =============================== */

.main-container {
  max-width: 1180px !important;
  font-family: "Segoe UI", Roboto, Arial, sans-serif;
  line-height: 1.65;
}

body {
  color: #243B53;
  background-color: #FAFBFC;
}

h1, h2, h3, h4 {
  font-weight: 750;
  color: #102A43;
}

h1.title {
  color: #102A43;
  font-size: 38px;
  font-weight: 800;
  margin-bottom: 8px;
}

h3.subtitle {
  color: #486581;
  font-weight: 500;
  margin-top: 0;
}

.author, .date {
  color: #627D98;
  font-size: 15px;
}

pre {
  border-radius: 12px;
  border: 1px solid #D9E2EC;
  background-color: #F8FAFC;
  padding: 14px;
}

code {
  color: #102A43;
}

.table {
  background-color: white;
}

blockquote {
  border-left: 5px solid #2F80ED;
  background-color: #F4F8FF;
  padding: 14px 18px;
  border-radius: 10px;
  color: #243B53;
}

.profile-card {
  background: linear-gradient(135deg, #102A43 0%, #243B53 56%, #334E68 100%);
  color: white;
  padding: 30px 34px;
  border-radius: 18px;
  margin: 22px 0 28px 0;
  box-shadow: 0 14px 34px rgba(16, 42, 67, 0.18);
}

.profile-card h2 {
  color: white;
  margin-top: 0;
  margin-bottom: 8px;
  font-size: 28px;
}

.profile-card p {
  color: #E6F6FF;
  margin: 5px 0;
  font-size: 15.5px;
}

.badge {
  display: inline-block;
  padding: 7px 12px;
  margin: 5px 7px 5px 0;
  border-radius: 999px;
  background-color: #D9E2EC;
  color: #102A43;
  font-size: 13px;
  font-weight: 700;
}

.badge-blue {
  background-color: #DCEEFB;
  color: #0B69A3;
}

.badge-green {
  background-color: #E3FCEC;
  color: #147D64;
}

.badge-gold {
  background-color: #FFF3C4;
  color: #8D2B0B;
}

.callout {
  border-left: 6px solid #2F80ED;
  background: #F4F8FF;
  padding: 17px 20px;
  border-radius: 12px;
  margin: 22px 0;
  box-shadow: 0 4px 12px rgba(47, 128, 237, 0.06);
}

.callout-success {
  border-left-color: #27AE60;
  background: #F4FFF8;
}

.callout-warning {
  border-left-color: #F2994A;
  background: #FFF8EF;
}

.callout-danger {
  border-left-color: #D64545;
  background: #FFF5F5;
}

.section-note {
  background: white;
  border: 1px solid #E5EAF0;
  border-radius: 14px;
  padding: 18px 20px;
  margin: 18px 0 24px 0;
  box-shadow: 0 8px 18px rgba(16, 42, 67, 0.05);
}

.metric-grid {
  display: grid;
  grid-template-columns: repeat(4, minmax(140px, 1fr));
  gap: 14px;
  margin: 20px 0 22px 0;
}

.metric-card {
  background: white;
  border: 1px solid #E5EAF0;
  border-radius: 16px;
  padding: 18px 16px;
  box-shadow: 0 8px 18px rgba(16, 42, 67, 0.06);
}

.metric-card .label {
  color: #627D98;
  font-size: 13px;
  font-weight: 700;
  text-transform: uppercase;
  letter-spacing: 0.04em;
}

.metric-card .value {
  color: #102A43;
  font-size: 28px;
  font-weight: 800;
  margin-top: 6px;
}

.caption-text {
  color: #627D98;
  font-size: 13px;
  font-style: italic;
  text-align: center;
  margin-top: -4px;
}

.footer-note {
  color: #627D98;
  font-size: 13px;
  border-top: 1px solid #D9E2EC;
  margin-top: 38px;
  padding-top: 18px;
}

@media print {
  .profile-card {
    box-shadow: none;
  }
  .metric-card {
    box-shadow: none;
  }
}
</style>

<div class="profile-card">
  <h2>Eliya Christopher Nandi</h2>
  <p><strong>Profession:</strong> Data Scientist | AI & Machine Learning Enthusiast</p>
  <p><strong>Email:</strong> eliyachristopher.nandi@community.unipa.it</p>
  <p><strong>Project:</strong> Statistical Machine Learning Portfolio Report</p>
  <p><strong>Focus:</strong> Country segmentation, dimensionality reduction, unsupervised learning, and evidence-based aid prioritization</p>
</div>

# Executive Summary

This report presents a professional end-to-end **unsupervised machine learning analysis** of country-level socio-economic and health indicators. The goal is to segment countries into meaningful groups and identify countries that may require urgent development support.

The workflow applies **feature engineering**, **exploratory data analysis**, **standardization**, **Principal Component Analysis (PCA)**, **K-means clustering**, **hierarchical clustering**, **silhouette validation**, and **need-score ranking**.

<span class="badge badge-blue">Data Science</span>
<span class="badge badge-blue">Statistical Machine Learning</span>
<span class="badge badge-green">PCA</span>
<span class="badge badge-green">K-means</span>
<span class="badge badge-green">Hierarchical Clustering</span>
<span class="badge badge-gold">Aid Prioritization</span>

<div class="callout">
<strong>Core Question</strong><br>
How can countries be grouped based on socio-economic and health indicators, and which countries should be prioritized for aid based on high child mortality, low GDP per capita, and low health expenditure?
</div>

# 1. Project Objectives

This analysis was designed to answer four practical questions:

1. What patterns exist among countries based on socio-economic and health indicators?
2. Can dimensionality reduction simplify the structure of the dataset while preserving most information?
3. How do K-means and hierarchical clustering compare for country segmentation?
4. Which countries show the greatest need for aid based on a transparent need-score framework?

# 2. Methodological Roadmap

The project follows a structured data science workflow:

1. Load and inspect the dataset.
2. Convert percentage-based economic indicators into actual values.
3. Conduct exploratory data analysis and outlier detection.
4. Standardize numeric variables for fair comparison.
5. Use PCA to reduce dimensionality.
6. Apply hierarchical clustering and K-means clustering.
7. Evaluate clustering quality using silhouette scores.
8. Profile clusters using meaningful socio-economic indicators.
9. Generate a need score and rank countries for aid prioritization.

<div class="section-note">
<strong>Why this structure matters:</strong><br>
A professional machine learning report should not only run models. It should explain why each method is used, what each code block does, and how the outputs support the final recommendation.
</div>

# 3. Load Required Libraries

Before running the analysis, the required R packages are loaded. These libraries support data manipulation, visualization, clustering, PCA, correlation analysis, and report formatting.

```{r load-libraries}
load_package <- function(package_name) {
  if (!requireNamespace(package_name, quietly = TRUE)) {
    stop(
      paste0(
        "Package '", package_name, "' is required but not installed. ",
        "Please install it using install.packages('", package_name, "')."
      )
    )
  }
  suppressPackageStartupMessages(
    library(package_name, character.only = TRUE)
  )
}

required_packages <- c(
  "tidyverse",
  "skimr",
  "ggcorrplot",
  "factoextra",
  "cluster",
  "knitr"
)

invisible(lapply(required_packages, load_package))

theme_set(theme_minimal(base_size = 12))
```

**Output interpretation:**  
If this code runs without errors, all required packages are available and the analysis environment is ready.

# 4. Load and Inspect the Dataset

This section imports the country dataset and performs an initial inspection. The goal is to understand the number of countries, variables, and the basic structure of the data before modeling.

```{r load-data}
if (!file.exists(params$data_path)) {
  stop(
    paste0(
      "Dataset not found: ", params$data_path,
      ". Please place Country-data.csv in the same folder as this R Markdown file."
    )
  )
}

df_raw <- read.csv(params$data_path)

n_countries <- nrow(df_raw)
n_variables <- ncol(df_raw)

cat("Number of countries:", n_countries, "\n")
cat("Number of variables:", n_variables, "\n")

head(df_raw) |>
  knitr::kable(caption = "Table 1. Preview of the Original Dataset")
```

**Output interpretation:**  
The preview confirms that each row represents a country and each column represents a socio-economic or health indicator.

```{r inspect-structure}
str(df_raw)
```

**Output interpretation:**  
The structure output shows variable types. The `country` variable is a character field, while the remaining variables are numeric indicators used for analysis.

# 5. Data Dictionary

The dataset includes a mix of demographic, health, and economic indicators. Understanding these variables is important before applying unsupervised learning.

```{r data-dictionary}
data_dictionary <- tibble::tribble(
  ~Variable, ~Description,
  "country", "Country name",
  "child_mort", "Child mortality rate",
  "exports", "Exports as percentage of GDP in the raw dataset",
  "health", "Health expenditure as percentage of GDP in the raw dataset",
  "imports", "Imports as percentage of GDP in the raw dataset",
  "income", "Net income per person",
  "inflation", "Annual inflation rate",
  "life_expec", "Average life expectancy",
  "total_fer", "Total fertility rate",
  "gdpp", "GDP per capita"
)

data_dictionary |>
  knitr::kable(caption = "Table 2. Data Dictionary")
```

**Output interpretation:**  
The variables provide enough information to compare countries across health outcomes, economic strength, trade activity, and population characteristics.

# 6. Feature Engineering

The original `exports`, `health`, and `imports` variables are percentages of GDP. This code converts them into actual monetary values by multiplying each percentage by GDP per capita. This makes the indicators easier to compare across countries.

```{r feature-engineering}
df <- df_raw |>
  mutate(
    exports = (exports * gdpp) / 100,
    health  = (health  * gdpp) / 100,
    imports = (imports * gdpp) / 100,
    gdpp    = as.numeric(gdpp),
    income  = as.numeric(income),
    country = factor(country)
  )

missing_values <- sum(sapply(df, function(x) sum(is.na(x))))

cat("Total missing values:", missing_values, "\n")

head(df) |>
  knitr::kable(
    digits = 2,
    caption = "Table 3. Dataset After Feature Engineering"
  )
```

**Output interpretation:**  
The converted values represent estimated actual exports, health expenditure, and imports per person. The missing-value check verifies whether imputation or row removal is required.

# 7. Dataset Overview After Cleaning

This section summarizes the cleaned dataset and gives a high-level view of the numeric variables.

```{r dataset-overview}
X <- df |> select(-country)

overview_tbl <- tibble(
  Metric = c("Countries", "Numeric indicators", "Missing values"),
  Value = c(nrow(df), ncol(X), missing_values)
)

overview_tbl |>
  knitr::kable(caption = "Table 4. Cleaned Dataset Overview")
```

```{r skim-summary}
skimr::skim(X)
```

**Output interpretation:**  
The summary statistics show central tendency, spread, and distribution characteristics. Large differences in scale across variables confirm the need for standardization before PCA and clustering.

# 8. Exploratory Data Analysis

Exploratory data analysis helps identify distributions, skewness, and potential outliers. This is essential before applying clustering because extreme values and scale differences can strongly influence unsupervised algorithms.

```{r prepare-eda}
X_long <- df |>
  select(country, all_of(names(X))) |>
  pivot_longer(
    cols = names(X),
    names_to = "feature",
    values_to = "value"
  )
```

**Output interpretation:**  
The dataset is reshaped into long format to make feature-by-feature visualization easier.

## 8.1 Distribution of Indicators

The following code plots the distribution of each numeric indicator. This helps identify skewed variables and unusual patterns.

```{r distribution-plot}
ggplot(X_long, aes(x = value)) +
  geom_histogram(
    bins = 25,
    fill = "#2F80ED",
    color = "white",
    alpha = 0.88
  ) +
  facet_wrap(~ feature, scales = "free") +
  labs(
    title = "Distribution of Socio-Economic and Health Indicators",
    subtitle = "Raw-scale distributions across all countries",
    x = "Value",
    y = "Number of Countries"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98"),
    strip.text = element_text(face = "bold", color = "#102A43")
  )
```

**Output interpretation:**  
Several variables are right-skewed, especially GDP per capita, income, exports, imports, and health expenditure. This is common in cross-country data because a small number of countries have much higher economic values than the majority.

## 8.2 Log-Scale View of Skewed Variables

Because economic variables often contain extreme high values, this code visualizes selected skewed variables on a log scale.

```{r log-scale-plot}
skew_vars <- c("gdpp", "income", "child_mort")

X_long |>
  filter(feature %in% skew_vars) |>
  ggplot(aes(x = value)) +
  geom_histogram(
    bins = 25,
    fill = "#56CCF2",
    color = "white",
    alpha = 0.9
  ) +
  facet_wrap(~ feature, scales = "free") +
  scale_x_continuous(trans = "log10") +
  labs(
    title = "Log10 View of Selected Skewed Variables",
    subtitle = "Log scaling improves visibility of heavily skewed distributions",
    x = "Log-scaled value",
    y = "Number of Countries"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98"),
    strip.text = element_text(face = "bold", color = "#102A43")
  )
```

**Output interpretation:**  
The log-scale view makes the spread of values easier to interpret and confirms that countries differ substantially in economic and health conditions.

## 8.3 Boxplots for Outlier Detection

This code uses boxplots to detect feature-level outliers. Outliers are important because they can influence PCA, K-means, and hierarchical clustering.

```{r boxplot-outliers}
ggplot(X_long, aes(x = feature, y = value, fill = feature)) +
  geom_boxplot(
    outlier.colour = "#D62728",
    outlier.size = 2.8,
    outlier.alpha = 0.75
  ) +
  coord_flip() +
  labs(
    title = "Boxplots for Outlier Detection",
    subtitle = "Red points represent potential outlying observations",
    x = "Feature",
    y = "Value"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98")
  )
```

**Output interpretation:**  
The plot highlights substantial outliers in GDP, income, exports, imports, and health expenditure. These outliers may represent highly developed or highly trade-dependent economies.

# 9. Standardization and Multivariate Outliers

Clustering requires variables to be on a comparable scale. This code standardizes all numeric variables using Z-score normalization.

```{r standardization}
X_scaled <- scale(X)

normalization_check <- tibble(
  feature = colnames(X_scaled),
  mean_after_scaling = round(colMeans(X_scaled), 4),
  sd_after_scaling = round(apply(X_scaled, 2, stats::sd), 4)
)

normalization_check |>
  knitr::kable(caption = "Table 5. Standardization Check")
```

**Output interpretation:**  
A mean close to 0 and standard deviation close to 1 confirm successful standardization.

## 9.1 Multivariate Outlier Ranking

This code calculates a multivariate outlier score based on each country's Euclidean distance from the standardized center of the dataset.

```{r multivariate-outliers}
outlier_rank <- tibble(
  country = df$country,
  outlier_score = sqrt(rowSums(X_scaled^2))
) |>
  arrange(desc(outlier_score))

head(outlier_rank, 10) |>
  knitr::kable(
    digits = 3,
    caption = "Table 6. Top 10 Multivariate Outliers Based on Z-score Distance"
  )
```

**Output interpretation:**  
The largest outlier scores identify countries whose overall socio-economic profiles differ most strongly from the rest of the dataset.

```{r outlier-barplot}
top_outliers <- outlier_rank |> slice_head(n = 10)

ggplot(top_outliers, aes(x = reorder(country, outlier_score), y = outlier_score, fill = outlier_score)) +
  geom_col(color = "#102A43", width = 0.72) +
  coord_flip() +
  scale_fill_gradient(low = "#56CCF2", high = "#D62728") +
  labs(
    title = "Top 10 Multivariate Outliers",
    subtitle = "Countries ranked by standardized multivariate distance",
    x = "Country",
    y = "Outlier Score"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98")
  )
```

**Output interpretation:**  
High-income and highly trade-oriented economies may appear as outliers because their GDP, income, export, import, or health expenditure values are far above the global average.

# 10. Correlation Analysis

Correlation analysis examines relationships between variables. Strong correlations suggest that dimensionality reduction can summarize the data efficiently.

```{r correlation-heatmap}
corr <- cor(X, use = "pairwise.complete.obs")

ggcorrplot(
  corr,
  type = "lower",
  lab = TRUE,
  lab_size = 3,
  colors = c("#D62728", "white", "#2F80ED")
) +
  labs(
    title = "Correlation Heatmap of Socio-Economic Indicators",
    subtitle = "Blue indicates positive correlation; red indicates negative correlation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98")
  )
```

**Output interpretation:**  
Variables such as GDP per capita, income, health expenditure, and life expectancy tend to capture development-related patterns. Child mortality and fertility often move in the opposite direction from development indicators.

# 11. Dimensionality Reduction with PCA

PCA reduces the original variables into a smaller set of uncorrelated components. This is useful because clustering in a lower-dimensional space can reduce noise and improve interpretability.

```{r pca-model}
pca_result <- prcomp(X_scaled)

explained_variance <- pca_result$sdev^2 / sum(pca_result$sdev^2)
cumulative_variance <- cumsum(explained_variance)

pca_variance_tbl <- tibble(
  component = paste0("PC", seq_along(explained_variance)),
  explained_variance_percent = round(explained_variance * 100, 2),
  cumulative_variance_percent = round(cumulative_variance * 100, 2)
)

num_components <- which(cumulative_variance >= 0.90)[1]

pca_variance_tbl |>
  knitr::kable(caption = "Table 7. PCA Explained Variance")
```

**Output interpretation:**  
This table shows how much information each principal component retains. Components are selected until at least 90% of total variance is retained.

```{r pca-components-summary}
cat("Number of principal components required to retain at least 90% variance:", num_components, "\n")
cat("Cumulative variance retained:", round(cumulative_variance[num_components] * 100, 2), "%\n")
```

**Output interpretation:**  
The selected components preserve most of the information in the dataset while reducing the number of dimensions used for clustering.

## 11.1 PCA Scree Plot

This code visualizes the contribution of each principal component.

```{r pca-scree-plot}
factoextra::fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 65),
  barfill = "#2F80ED",
  barcolor = "#2F80ED",
  main = "PCA Scree Plot"
) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#102A43")
  )
```

**Output interpretation:**  
The first few components explain most of the variance, which justifies using PCA before clustering.

## 11.2 Cumulative Explained Variance

This code plots cumulative explained variance and highlights the 90% threshold.

```{r cumulative-variance-plot}
cumulative_df <- tibble(
  component_number = seq_along(cumulative_variance),
  cumulative_variance = cumulative_variance
)

ggplot(cumulative_df, aes(x = component_number, y = cumulative_variance)) +
  geom_line(color = "#2F80ED", linewidth = 1.1) +
  geom_point(color = "#D62728", size = 3) +
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "#27AE60", linewidth = 1) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = cumulative_df$component_number) +
  labs(
    title = "Cumulative Explained Variance",
    subtitle = "Dashed line indicates the 90% variance threshold",
    x = "Number of Principal Components",
    y = "Cumulative Explained Variance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98")
  )
```

**Output interpretation:**  
The curve shows that a small number of components can summarize most of the dataset.

<div class="callout callout-success">
<strong>PCA Decision:</strong><br>
The analysis retains <strong>`r num_components` principal components</strong>, preserving approximately <strong>`r round(cumulative_variance[num_components] * 100, 2)`%</strong> of the total variance.
</div>

# 12. Optional Autoencoder Comparison

An autoencoder is a neural-network-based dimensionality reduction method. It can capture non-linear patterns, but it requires additional deep learning dependencies. For a clean professional report, this section is optional.

To run it, change `run_autoencoder` from `false` to `true` in the YAML header.

```{r autoencoder, eval=params$run_autoencoder}
if (!requireNamespace("keras3", quietly = TRUE)) {
  stop("The keras3 package is required to run the autoencoder section.")
}

library(keras3)

input_dim <- ncol(X_scaled)
latent_dim <- num_components

input_layer <- layer_input(shape = c(input_dim))

encoder_hidden <- input_layer |>
  layer_dense(units = 128, activation = "relu")

encoder_output <- encoder_hidden |>
  layer_dense(units = latent_dim, activation = "relu")

decoder_hidden <- encoder_output |>
  layer_dense(units = 128, activation = "relu")

decoder_output <- decoder_hidden |>
  layer_dense(units = input_dim, activation = "linear")

autoencoder <- keras_model(inputs = input_layer, outputs = decoder_output)

autoencoder |>
  compile(
    optimizer = "adam",
    loss = "mean_squared_error"
  )

history <- autoencoder |>
  fit(
    X_scaled,
    X_scaled,
    epochs = 50,
    batch_size = 32,
    validation_split = 0.2,
    verbose = 0
  )

encoder <- keras_model(
  inputs = input_layer,
  outputs = encoder_output
)

encoded_data <- encoder |> predict(X_scaled)

encoded_df <- as.data.frame(encoded_data)
colnames(encoded_df) <- paste0("Latent_", seq_len(ncol(encoded_df)))

ggplot(encoded_df, aes(x = Latent_1, y = Latent_2)) +
  geom_point(alpha = 0.75, color = "#D62728", size = 2.6) +
  labs(
    title = "2D Latent Space from Autoencoder",
    x = "Latent Dimension 1",
    y = "Latent Dimension 2"
  ) +
  theme_minimal(base_size = 12)
```

**Output interpretation:**  
If enabled, the autoencoder output provides a non-linear compressed representation of the dataset. PCA is retained as the main method because it is interpretable, reproducible, and sufficient for the observed correlated structure.

# 13. Prepare PCA Data for Clustering

This code extracts the selected PCA components and prepares them for clustering.

```{r prepare-pca-clustering}
pca_components <- as.data.frame(pca_result$x[, 1:num_components])
pca_components$country <- df$country

pca_matrix <- pca_components |> select(-country)

head(pca_components) |>
  knitr::kable(
    digits = 3,
    caption = "Table 8. PCA Component Scores Used for Clustering"
  )
```

**Output interpretation:**  
Each country is now represented by a smaller set of PCA component scores instead of the original variables.

# 14. Hierarchical Clustering

Hierarchical clustering builds a tree-like structure of country similarity. Ward's method is selected because it tends to create compact and interpretable clusters.

```{r hierarchical-distance}
dist_pca <- dist(pca_matrix, method = "euclidean")
hc_ward <- hclust(dist_pca, method = "ward.D2")
```

**Output interpretation:**  
The Euclidean distance matrix captures similarity between countries in PCA space. Ward's method uses this distance matrix to create hierarchical clusters.

## 14.1 Dendrogram

This code visualizes the hierarchical clustering tree.

```{r hierarchical-dendrogram}
plot(
  hc_ward,
  labels = FALSE,
  hang = -1,
  main = "Hierarchical Clustering Dendrogram: Ward's Method",
  xlab = "",
  ylab = "Height"
)
```

**Output interpretation:**  
The dendrogram shows how countries merge into larger groups. Large vertical jumps indicate stronger separation between clusters.

## 14.2 Cut the Dendrogram into Clusters

This code cuts the dendrogram into `r params$k_clusters` clusters.

```{r cut-hierarchical}
k_final <- params$k_clusters

hc_clusters <- cutree(hc_ward, k = k_final)

pca_components$cluster_hc <- factor(hc_clusters)

table(pca_components$cluster_hc) |>
  knitr::kable(caption = "Table 9. Hierarchical Clustering: Cluster Sizes")
```

**Output interpretation:**  
The cluster-size table shows how many countries belong to each hierarchical cluster.

```{r highlighted-dendrogram}
plot(
  hc_ward,
  labels = FALSE,
  hang = -1,
  main = paste0("Hierarchical Clustering Dendrogram: k = ", k_final),
  xlab = "",
  ylab = "Height"
)

rect.hclust(hc_ward, k = k_final, border = 2:(k_final + 1))
```

**Output interpretation:**  
The colored rectangles show the selected cluster structure.

## 14.3 Hierarchical Clustering Silhouette Analysis

Silhouette analysis measures how well each country fits within its assigned cluster.

```{r hierarchical-silhouette}
sil_hc <- silhouette(hc_clusters, dist_pca)

factoextra::fviz_silhouette(sil_hc) +
  ggtitle(paste0("Silhouette Plot: Hierarchical Clustering, k = ", k_final)) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", color = "#102A43"))

avg_sil_hc <- mean(sil_hc[, 3])

cat("Average silhouette width for hierarchical clustering:", round(avg_sil_hc, 3), "\n")
```

**Output interpretation:**  
Higher silhouette values indicate better-defined clusters. Values closer to 1 indicate strong separation, while values near 0 indicate overlap.

# 15. K-means Clustering

K-means clustering partitions countries into a fixed number of groups by minimizing within-cluster variation.

## 15.1 Elbow Method

The elbow method helps inspect a reasonable number of clusters by plotting within-cluster sum of squares for different values of `k`.

```{r elbow-method}
wss <- sapply(1:10, function(k) {
  kmeans(pca_matrix, centers = k, nstart = 25)$tot.withinss
})

elbow_df <- tibble(k = 1:10, wss = wss)

ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line(color = "#2F80ED", linewidth = 1.1) +
  geom_point(color = "#D62728", size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Elbow Method for K-means",
    subtitle = "The elbow point suggests a practical number of clusters",
    x = "Number of Clusters",
    y = "Total Within-Cluster Sum of Squares"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98")
  )
```

**Output interpretation:**  
A visible bend in the curve suggests that adding more clusters after that point gives diminishing returns.

## 15.2 Fit K-means Model

This code fits K-means using the selected number of clusters.

```{r fit-kmeans}
set.seed(123)

kmeans_result <- kmeans(
  pca_matrix,
  centers = k_final,
  nstart = 25
)

kmeans_clusters <- kmeans_result$cluster

table(kmeans_clusters) |>
  knitr::kable(caption = "Table 10. K-means Clustering: Cluster Sizes")
```

**Output interpretation:**  
The table shows the number of countries assigned to each K-means cluster.

## 15.3 Visualize K-means Clusters

This code plots the K-means clusters in the first two PCA dimensions.

```{r kmeans-plot}
pca_cluster_kmeans <- tibble(
  country = df$country,
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  Cluster = factor(kmeans_clusters)
)

ggplot(pca_cluster_kmeans, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3, alpha = 0.82) +
  labs(
    title = "K-means Clusters in PCA Space",
    subtitle = "Countries projected onto the first two principal components",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98")
  )
```

**Output interpretation:**  
The plot provides a visual representation of country groups in reduced PCA space.

## 15.4 K-means Silhouette Analysis

This code evaluates the quality of K-means clustering using silhouette width.

```{r kmeans-silhouette}
sil_kmeans <- silhouette(kmeans_clusters, dist(pca_matrix))

factoextra::fviz_silhouette(sil_kmeans) +
  ggtitle(paste0("Silhouette Plot: K-means, k = ", k_final)) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", color = "#102A43"))

avg_sil_kmeans <- mean(sil_kmeans[, 3])

cat("Average silhouette width for K-means:", round(avg_sil_kmeans, 3), "\n")
```

**Output interpretation:**  
The average silhouette width summarizes overall cluster quality for K-means.

# 16. Comparison of Clustering Methods

This section compares K-means and hierarchical clustering using average silhouette width.

```{r method-comparison}
comparison_tbl <- tibble(
  Method = c("Hierarchical Clustering", "K-means Clustering"),
  Main_Idea = c(
    "Builds a nested tree of country similarity using Ward's method",
    "Partitions countries by minimizing within-cluster variation"
  ),
  Number_of_Clusters = c(k_final, k_final),
  Average_Silhouette_Width = c(avg_sil_hc, avg_sil_kmeans)
) |>
  mutate(Average_Silhouette_Width = round(Average_Silhouette_Width, 3))

comparison_tbl |>
  knitr::kable(caption = "Table 11. Clustering Method Comparison")
```

**Output interpretation:**  
The method with the higher average silhouette width provides stronger separation under this configuration. Both methods are retained because they offer complementary views of the country groupings.

<div class="callout">
<strong>Professional Interpretation:</strong><br>
Hierarchical clustering is useful for understanding nested similarity structures, while K-means provides a direct partitioning of countries into groups. Comparing both strengthens the reliability of the final aid-prioritization decision.
</div>

# 17. Cluster Profiling

Cluster profiling translates statistical clusters into meaningful socio-economic categories.

```{r cluster-profiling}
analysis_df <- df |>
  mutate(
    cluster_kmeans = factor(kmeans_clusters),
    cluster_hc = factor(hc_clusters)
  )

profile_clusters <- function(data, cluster_col) {
  data |>
    group_by({{ cluster_col }}) |>
    summarise(
      countries = n(),
      avg_gdpp = mean(gdpp, na.rm = TRUE),
      avg_income = mean(income, na.rm = TRUE),
      avg_exports = mean(exports, na.rm = TRUE),
      avg_health = mean(health, na.rm = TRUE),
      avg_child_mort = mean(child_mort, na.rm = TRUE),
      .groups = "drop"
    )
}

kmeans_profile <- profile_clusters(analysis_df, cluster_kmeans)
hc_profile <- profile_clusters(analysis_df, cluster_hc)

kmeans_profile |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  knitr::kable(caption = "Table 12. K-means Cluster Profile")

hc_profile |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  knitr::kable(caption = "Table 13. Hierarchical Cluster Profile")
```

**Output interpretation:**  
Clusters with high child mortality, low GDP per capita, and low health expenditure represent countries with more severe socio-economic and health challenges.

# 18. Identify Highest-Need Clusters

This code identifies the highest-need cluster in each clustering method. The ranking prioritizes high child mortality, low GDP per capita, and low health spending.

```{r priority-clusters}
priority_cluster_kmeans <- kmeans_profile |>
  arrange(desc(avg_child_mort), avg_gdpp, avg_health) |>
  slice(1)

priority_cluster_hc <- hc_profile |>
  arrange(desc(avg_child_mort), avg_gdpp, avg_health) |>
  slice(1)

priority_cluster_kmeans |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  knitr::kable(caption = "Table 14. Highest-Need Cluster Based on K-means")

priority_cluster_hc |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  knitr::kable(caption = "Table 15. Highest-Need Cluster Based on Hierarchical Clustering")
```

**Output interpretation:**  
The highest-need cluster is the group with the most severe combination of poor health and weak economic indicators.

# 19. Need Score Construction

A transparent need score is calculated to rank countries inside the priority group.

The score is defined as:

\[
\text{Need Score} = z(\text{child mortality}) - z(\text{GDP per capita}) - z(\text{health expenditure})
\]

Higher values indicate greater need.

```{r need-score}
priority_countries <- analysis_df |>
  filter(
    cluster_kmeans == priority_cluster_kmeans$cluster_kmeans |
      cluster_hc == priority_cluster_hc$cluster_hc
  ) |>
  mutate(
    normalized_child_mort = as.numeric(scale(child_mort)),
    normalized_gdpp = as.numeric(scale(gdpp)),
    normalized_health = as.numeric(scale(health)),
    need_score = normalized_child_mort - normalized_gdpp - normalized_health
  ) |>
  arrange(desc(need_score))

priority_countries |>
  select(country, child_mort, gdpp, income, health, need_score) |>
  mutate(need_score = round(need_score, 3)) |>
  head(15) |>
  knitr::kable(caption = "Table 16. Top 15 Countries Ranked by Need Score")
```

**Output interpretation:**  
Countries at the top of the table have the strongest evidence of need based on the selected indicators.

```{r need-score-plot}
priority_countries |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(country, need_score), y = need_score, fill = need_score)) +
  geom_col(show.legend = FALSE, width = 0.75) +
  coord_flip() +
  scale_fill_gradient(low = "#56CCF2", high = "#D62728") +
  labs(
    title = "Top 15 Countries Ranked by Need Score",
    subtitle = "Higher score indicates higher aid priority",
    x = "Country",
    y = "Need Score"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98")
  )
```

**Output interpretation:**  
The ranking provides a clear, explainable output that can support policy discussion and resource allocation planning.

# 20. Cluster-Level Indicator Comparison

The following plot compares major indicators across K-means clusters.

```{r cluster-boxplots}
analysis_df |>
  pivot_longer(
    cols = c(child_mort, gdpp, income, health),
    names_to = "indicator",
    values_to = "value"
  ) |>
  ggplot(aes(x = cluster_kmeans, y = value, fill = cluster_kmeans)) +
  geom_boxplot(alpha = 0.86) +
  facet_wrap(~ indicator, scales = "free_y") +
  labs(
    title = "Key Indicators by K-means Cluster",
    subtitle = "Comparing economic and health characteristics across country groups",
    x = "K-means Cluster",
    y = "Value"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", color = "#102A43"),
    plot.subtitle = element_text(color = "#627D98"),
    strip.text = element_text(face = "bold", color = "#102A43")
  )
```

**Output interpretation:**  
The boxplots show how clusters differ in child mortality, GDP per capita, income, and health expenditure. The most vulnerable cluster is expected to have high child mortality and lower economic/health-resource indicators.

# 21. Country Lists by Cluster

This section provides a transparent view of which countries belong to each cluster.

```{r countries-by-cluster}
countries_by_kmeans <- analysis_df |>
  arrange(cluster_kmeans, country) |>
  group_by(cluster_kmeans) |>
  summarise(
    countries = paste(as.character(country), collapse = ", "),
    .groups = "drop"
  )

countries_by_hc <- analysis_df |>
  arrange(cluster_hc, country) |>
  group_by(cluster_hc) |>
  summarise(
    countries = paste(as.character(country), collapse = ", "),
    .groups = "drop"
  )

countries_by_kmeans |>
  knitr::kable(caption = "Table 17. Countries by K-means Cluster")

countries_by_hc |>
  knitr::kable(caption = "Table 18. Countries by Hierarchical Cluster")
```

**Output interpretation:**  
This section improves interpretability by showing the country membership of each cluster instead of only presenting numerical summaries.

# 22. Final Recommendation

<div class="callout callout-success">
<strong>Recommendation:</strong><br>
Aid should be prioritized toward countries in the highest-need cluster, especially those with the highest need scores. These countries show the strongest combination of high child mortality, low GDP per capita, and low health expenditure.
</div>

The analysis suggests that aid programs should focus on:

1. Healthcare investment and child mortality reduction.
2. Economic development and poverty reduction.
3. Education and infrastructure development.
4. Country-specific validation using local governance, population, humanitarian access, and conflict-risk data.

# 23. Final Conclusion

This project demonstrates how unsupervised machine learning can support data-driven decision-making in international development and aid prioritization. PCA successfully reduced the dimensionality of the dataset while retaining most of the original information. K-means and hierarchical clustering both produced interpretable country groups. The need-score framework then translated clustering results into an actionable ranking of countries.

<div class="callout callout-warning">
<strong>Important Professional Note:</strong><br>
This analysis is intended for statistical learning, portfolio demonstration, and decision-support exploration. Real-world aid allocation should combine quantitative analysis with domain expertise, field-level information, population size, political context, humanitarian access, and ethical review.
</div>

# 24. Reproducibility Instructions

To reproduce this report:

1. Place `Country-data.csv` in the same folder as this `.Rmd` file.
2. Open the `.Rmd` file in RStudio.
3. Install missing packages if prompted.
4. Click **Knit** to generate the HTML report.
5. Upload both the `.Rmd` and knitted `.html` file to GitHub.

# 25. Session Information

The session information below records the R version and package environment used to generate the analysis.

```{r session-info}
sessionInfo()
```

<div class="footer-note">
Prepared by Eliya Christopher Nandi — Data Scientist | AI & Machine Learning Enthusiast.
</div>
